Making Sequential Programs Deterministic

### 2019-03-26 design.tex
polymorphic extensions

Figure~\ref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following annotation 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantion \codeid{public void add(NonDet ArrayList<E> this, NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another valiant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being resolved as \codeid{public void add(Det ArrayList<E> this, Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.

>>>>

Figure~\ref{fig-poly-resolution} presents the rule for type checking of methods annotated with \<PolyDet> or its variants.
A polymorphic method type checks if every instantiation of \<PolyDet> type checks. If this method has \<PolyDet>$\uparrow$ or \<PolyDet>$\downarrow$
in its signature then the instantiation of \<PolyDet> at that location gets replaced by the appropriate type as shown in Table~\ref{tab-poly-resolutions}.

@@

In order to provide soundness guarantees, we must ensure that a collection that is
declared as \<Det> is not side-effected in non-deterministic ways. Consider the following annotation 
on the List add method: \codeid{public void add(PolyDet ArrayList<E> this, PolyDet int index, E element)}.
Suppose a client calls this method as follows:
\begin{verbatim}
@Det ArrayList<@Det String> list;
@NonDet int random;
@Det String str;
list.add(random, str);
\end{verbatim}
This will result in the instantion \codeid{public void add(NonDet ArrayList<E> this, NonDet int index, E element)} and
the method invocation will type check. This is problematic because even though \<list> was declared to be \<Det>,
the type checker allows the addition of an element at a \<NonDet> location violating our determinism guarantees.
We prevent this behavior by introducing another variant of \<PolyDet>, \<PolyDet("use")> that does not affect
type instantiations. This is especially useful in preventing methods from non-deterministically modifying the state
of a deterministic receiver. To have this effect, the library method \<ArrayList.add> is annotated as:
\codeid{public void add(PolyDet ArrayList<E> this, PolyDet("use") int index, E element)}.
Since \<PolyDet("use")> has no effect on type instantiations, calling this method on a \<Det List>
will result in the method declaration being resolved as \codeid{public void add(Det ArrayList<E> this, Det int index, E element)}
irrespective of the type of \<index>, thereby preventing undesired side-effects.
